---
title: "Pet Lakes Example"
author: "Amalia Handler"
date: "2024-01-06"
output: html_document
editor_options: 
  chunk_output_type: console
---

Purpose of this script is to compile a list of commonly studied or identified mountain lakes across the US, run these lakes through the mountain lake identification process, and assess which of the lakes are deemed "mountain lakes" via the process I developed.

```{r load packages, echo = F, include = F, message = F, warning = F}
library(dplyr)
library(sf)
library(mapview)
library(nhdplusTools)
library(DT)
library(here)
```


List compiled with the help of Lara Jansen

###### List of pet mountain lakes
* Castel lake, CA
* Lake Tahoe, CA
* Crater Lake, OR
* Loch Vale watershed (CO), including:
**   The Loch
**   Sky Pond
* Flathead Lake, MT
* Bear Tooth Lake, WY
* Lake Placid, NY
* Lake George, NY
* Lake Champlain, NY
* Lake Powell, AZ
* Lake Winnippesaukee, NH
* Lake Sunapee, NH
* Lake Coeur d'Alene, ID
* Lake Pend Oreille, ID
* Mountain Lake, VA
* Hoh Lake, Olympics WA
* Emerald Lake, CA
* Mowich Lake, Mt Raineir, WA
* Upper Carrol Lake, UT
* Mirror Lake, NH (Hubbard Brook)

NHDPlus waterbody COMIDs were retrieved by searching lake name in the WATER GeoViewer
https://www.epa.gov/waterdata/waters-geoviewer

```{r get lake boundaries, echo = F, include = F}
# Spatial data directory
dir <- "O:/PRIV/CPHEA/PESD/COR/CORFILES/Geospatial_Library_Projects/AmaliaHandler/Mountain Lakes 2022-2025 Amalia Handler 1035b/"

# Retrieve waterbodies from NHDPlus
pet_in <- readr::read_csv("./inst/data/Pet_lake_comids.csv")

pets <- get_waterbodies(id = pet_in$COMID)

# Note that the name Harrison Slough is returned for Coeur d’Alene Lake

# Read in LakeCat NLCD table to get the IDs of lakes that are in lakecat
nlcd_lakecat <- readr::read_csv('O:/PRIV/CPHEA/PESD/COR/CORFILES/Geospatial_Library_Projects/LakeCat/FTP_Staging/FinalTables/NLCD2011.csv')

# Which of the pet mountain lakes are included in LakeCat?

# Add a column identifying the lakes in LakeCat
lakecat_table <- mutate(nlcd_lakecat, in_lakecat = TRUE)

# Crosswalk NLA sample with LakeCat
pets |>
  st_drop_geometry() |>
  left_join(select(lakecat_table, COMID, in_lakecat), by = c("comid" = "COMID")) |>
  mutate(in_lakecat = ifelse(is.na(in_lakecat), FALSE, in_lakecat)) |>
  select(comid, in_lakecat)

# Good news! They are all in LakeCat!
```


```{r compile catchments, echo = F, eval = F}

# LakeCat off-network catchments
off_net_cats <- read_sf('O:/PRIV/CPHEA/PESD/COR/CORFILES/Geospatial_Library_Projects/LakeCat/FTP_Staging/Framework/LkCat_Frame_min/shps/allBasins.shp')

pet_off_net <- filter(off_net_cats, COMID %in% pet_in$COMID)

# On-network 
# Compiled from StreamCat

# COMIDs for the on-network lakes
pet_on_net <- pet_in$COMID[!pet_in$COMID %in% pet_off_net$COMID]

# Filter to lakes in the example ecoregions that are presumed on the network
on_net <- lakecat_table |>
  filter(COMID %in% pet_on_net) |>
  select(COMID, inStreamCat)

# Check how many lakes this identified. Will be less than those in the NHD due to some size dependencies.
table(on_net$inStreamCat, exclude = NULL)

# Get the corresponding catchment IDs for lakes
lookup_dir <- 'O:/PRIV/CPHEA/PESD/COR/CORFILES/Geospatial_Library_Projects/LakeCat/LakeCat_Framework/joinTables'

lookup <- list.files(lookup_dir, full.names = TRUE) |>  
    purrr::map_dfr(readr::read_csv)

# Only keep network COMIDs that have a corresponding lake WBAREACOMID
on_net_wbcomid <- filter(lookup, wbCOMID %in% on_net$COMID)

# Gather the NHDPlus catchments for these lakes

# Read in catchments and flowlines from NHDPlusV21
loc_gdb <- 'O:/PRIV/CPHEA/PESD/COR/CORFILES/Geospatial_Library_Resource/Physical/HYDROLOGY/NHDPlusV21/NHDPlusNationalData/NHDPlusV21_National_Seamless_Flattened_Lower48.gdb'

# These will take a long time to read in
cats       <- read_sf(dsn = loc_gdb, layer = 'Catchment')
flowlines  <- read_sf(dsn = loc_gdb, layer = 'NHDFlowline_Network')

# Filter to the flowlines and catchments that are on-network and in the example ecoregions
flow_ex_lakes <- filter(flowlines, WBAREACOMI %in% on_net_wbcomid$wbCOMID)

cats$wbCOMID <- lookup$wbCOMID[match(cats$FEATUREID, lookup$catCOMID)]
cats_filt    <- filter(cats, FEATUREID %in% flow_ex_lakes$COMID)

cats_filt$wbCOMID <- flow_ex_lakes$WBAREACOMI[match(cats_filt$FEATUREID, flow_ex_lakes$COMID)]

# Resolve invalid geometry
cats_valid <- cats_filt  |>  
  st_make_valid(cats_filt) |>
  group_by(wbCOMID)  |> 
  summarise(AreaSqKM = sum(AreaSqKM))

# Check that the wbCOMIDs are the same as the lake COMIDs
on_net$COMID[!on_net$COMID %in% cats_valid$wbCOMID]

# Marry the on and off network catchments, then write out a shapefile
clean_offnet <- pet_off_net |>
  select(COMID, AreaSqKM, geometry) |>
  rename(wbCOMID = COMID) |>
  mutate(on_network = FALSE, .after = wbCOMID)

clean_onnet <- cats_valid |>
  mutate(on_network = TRUE, .after = wbCOMID) |>
  rename(geometry = Shape) |>
  st_transform(st_crs(clean_offnet))

cats_combine <- rename(bind_rows(clean_offnet, clean_onnet), COMID = wbCOMID)

# Write the combined file
st_write(cats_combine, paste0(dir, "ExampleMtnLakes.gpkg"), layer = "ExampleMtnLakeCatchments")

# Write out the lake geometry
st_write(pets, dsn = paste0(dir, "ExampleMtnLakes.gpkg"), layer = "ExampleMtnLakePolygons"))

# Use the above file to run the IdentifyMountainLakes.Rmd script

# Write the file with just the catchment areas to the data location on the C drive
readr::write_csv(st_drop_geometry(cats_combine), "/inst/data/Pet_lake_catchment_area.csv")

```


```{r compile processed data, echo = F, eval = F, include = F}
# Read file names
files <- list.files(paste0(data_dir, "/data/pet_mountain_lakes"), pattern = "lake_", full.names = T)

# Create a master df with all data
pet_lakes <- do.call(rbind, lapply(files, readRDS))

# Write the data to csv for analysis
readr::write_csv(pet_lakes, "./inst/data/Pet_mountain_lakes_output.csv") 
readr::write_csv(pet_lakes, "./inst/data/Pet_mountain_lakes_polygons.csv") 
```


How much of the pet mountain lake catchments are designated as "mountainous" according to the landforms file?

```{r, echo = F, message=F, warning=F}
# Load the process for the pet lakes catchments (note these include the lake area)
pet_catch <- readr::read_csv(paste0(here(), "/inst/data/Pet_mountain_lakes_output.csv"))

pet_input <- readr::read_csv(paste0(here(), "/inst/data/Pet_lake_comids.csv")) |>
  mutate(name = ifelse(COMID == 24381227, "Coeur d'Alene Lake", name),
         range = ifelse(COMID == 24381227, "Shelkirk & Coeur d'Alene", range))

catch_area <- readr::read_csv(paste0(here(), "/inst/data/Pet_lake_catchment_area.csv"))

# Output of the process for just the pet mountain lake polygons (does not include any of the catchment area)
pet_lakes <- readr::read_csv(paste0(here(), "/inst/data/Pet_mountain_lakes_polygons.csv"))

# Proportion of the catchment that is mountain area
# pet_lakes <- pet_catch |>
#   mutate(percent_mtn_area = round(((low_mtn_pix + high_mtn_pix)/tot_crop_pix) * 100, 0))

# Now removed the lake area from the catchment area (note that the two dataframes are ordered the same)
mtn_pix_no_lake <- (pet_catch$low_mtn_pix + pet_catch$high_mtn_pix) - (pet_lakes$low_mtn_pix + pet_lakes$high_mtn_pix)
tot_pix_no_lake <- pet_catch$tot_crop_pix - pet_lakes$tot_crop_pix
pet_no_lake <- tibble(comid = pet_lakes$unique_id, mtn_pix = mtn_pix_no_lake, tot_pix = tot_pix_no_lake)

# Join with lake names
pet <- pet_no_lake |>
  mutate(pct_mtn_area = round(mtn_pix/tot_pix * 100), 0) |>
  select(comid, pct_mtn_area) |>
  left_join(pet_input, by = c("comid" = "COMID")) |>
  arrange(range) |>
  select(-comid) |>
  relocate(name, state, range, pct_mtn_area, citation)


datatable(pets, caption = htmltools::tags$caption("Pet Mountain Lakes", style='color:black; font-size: 20px; text-align:left')) %>%
  DT::formatStyle(columns = names(pet), fontSize = "75%")

usethis::use_data(pet)

```

